Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2018-12-05 article_springer/article_springer.Rmd

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T_\ell \sim {\rm ML}(\beta, k^{1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not 
memoryless, and the probability density of the time $t$ until the next threshold 
crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{1/\beta} \sigma_0)$ is the probability density of 
${\rm ML}(\beta, k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability 
distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). 
The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a 
decreasing function for the Mittag-Leffler distribution. 
```{r hazard, fig.height=5, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

By the Theorem, the distribution 
of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$.
The distribution of $J_k$ is irrelevant for the inference on $\beta$ and 
$\sigma_0$ (we have chosen unit exponential random variables). 
Figure \ref{fig:sim} displays plots of $\hat \beta(\ell)$ and 
$\hat \sigma(\ell)$ vs. $k$; recall that $k$ is the index of the order
statistics of $J_k$ at which the threshold $\ell$ is placed. 
Dotted lines show 95% confidence intervals, which are derived from the 
asymptotic normality of the log-moments estimators [@Cahoy2013] and the $\delta$-method [@MittagLeffleR].
The dashed lines show the actual values of $\beta$ resp.\ $\sigma_0$, 
showing that our inference method identifies the 
parameters correctly.

@@

```{r flare-diagnostics-2, fig.height=4.5, fig.width=8, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(1,2))
empcopula(flares_ctre)
flares_ctre %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.8, main = "Mittag-Leffler QQ Plot")
```

@@

>>>>

Since the exponential distribution is nested in the Mittag-Leffler family of distributions, an appropriate way to choose between a model with exponential and Mittag-Leffler inter-exceedance times seems to be a Likelihood ratio test. Although the two models are nested, the assymptotic distribution is not $\chi^2_1$-distributed, the classical Theorem of Wilk doesn't apply since under $H_0$ the parameter $\beta$ of the Mittag-Leffler distribution is equal to one and hence lies under $H_0$ on the boundary of the parameter space $(0,1]$. Nonetheless, one can perform a bootstrapped log-likelihood ratio test. In Figure \ref{Fig:LRT} one can see the power for the bootstrapped LRT for Mittag-Leffler distributions with different tail parameters. As expected the power gets smaller for tail parameters close to one, since the Mittag-Leffler distribution converges for $\beta \rightarrow 1$ to an exponential distribution. Consequently, for tail parameters close to one it is hard to differentiate a Mittag-Lefller distribution from an exponential.

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T(\ell) \sim {\rm ML}(\beta, k^{-1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not memoryless, and the probability density of the time $t$ until the next threshold crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{-1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{-1/\beta} \sigma_0)$ is the probability density of ${\rm ML}(\beta, -k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{-1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{-1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a decreasing function for the Mittag-Leffler distribution. 
```{r hazard, fig.height=4, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

By the Theorem, the distribution 
of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$. we have chosen $\beta=1$ in the stable as well in the Pareto case. Figure \ref{Ex1figs} displays plots of $\hat \beta(k)$ vs. $k$ for the first example with stable distributed waiting times. In the left picture we used the log moment estimator, in the middle picture maximum likelihood estimates (both applied to the inter-exceedance times) and for the right picture hill estimators applied to the inter-arrival times. Recall that $k$ is the index of the order statistics of $J_k$ at which the threshold $\ell$ is placed.

```{r Ex1figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex1figs} Stability Plots for m=100 simulation runs for Example 1 (stable distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Tail08_mles.pdf","SimuStudy_StabilityPlots_Tail08_hill.pdf"))
```
Figure \ref{Ex2figs} shows the estimates fordifferent $k$'s and for the different estimators for the tail parameter $\beta$. Again one can see, that there is much more variance between the stability plots for the different simulation runs in case of the hill estimator compared to the other two estimators.  
```{r Ex2figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex2figs} Stability Plots for m=100 simulation runs for Example 2 (Pareto distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Ex2_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_mles.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_hill.pdf"))
```
To show how the different estimates behave in the case that $\beta=1$, hence, the inter-exceedance times are exponetial distributed, we simulated in a last example exponetial distributed waiting times.

```{r Ex3figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex3figs} Stability Plots for m=100 simulation runs for Example 3 (Exponentially distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Ex3_logMoms.pdf","SimuStudy_StabilityPlots_Ex3_mles.pdf","SimuStudy_StabilityPlots_Ex3_hill_times.pdf"))
```

@@

```{r flare-diagnostics-2, fig.height=4.5, fig.width=8, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(1,2))
empcopula(flares_ctre)
flares_ctre %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.85, main = "Mittag-Leffler QQ Plot")
```

Figure \ref{fig:flares} shows the stability plots for the solar flare data, on the left for the tail parameter and on the right for the scale parameter. Dotted lines show 95% confidence intervals, which are derived from the asymptotic normality of the log-moments estimators [@Cahoy2013] and the $\delta$-method [@MittagLeffleR, dashed lines show the actual values of $\beta$ resp.\ $\sigma_0$. The stability plot for the tail stabilizes nicely round about 0.85 (dotted line), while it is a little bit harder to deduce an estimate for the scale parameter. Howerever, the stability plot for the scale parameter seems to stabilize round about $3 \times 10^7$.

@@

The fit with a Mittag-Leffler distribution ($\beta = 0.85$) is good, though there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot. We also counduct a bootstrapped LRT for the null hypothesis of exponential distributed inter-arrival times and received a $p$-value of $p<0.01$.

```{r likelihood-ratio-test}
library("MittagLeffleR")
library("stats")

# ML_LRT_boot<-function(data,signi=0.05,m=1000,n){
# if(is.null(dim(data))){
# mle_data<-mlmle(data)                                                   #Calculate loglikelihoods for obserevd data
# teststat<-2*(mle_data$loglik-mle_data$null_loglik)}                     #Calculate teststatistic for obserevd data
# else{
# l<-apply(data,MARGIN=2,FUN=mlmle)                                       #Extension to multivariate datasets for simulation studies
# teststat<-2*(unlist(lapply(l,"[[",2))-unlist(lapply(l,"[[",3)))
# }
# 
# x<-replicate(m,rexp(n,rate=1))                                          #Sample data under Null
# l2<-apply(x,MARGIN=2,FUN=mlmle)                                         #Calculate loglikelihoods for each simulated data set
# test_stat_null<-2*(unlist(lapply(l2,"[[",2))-unlist(lapply(l2,"[[",3))) #Calculate teststatistic for each simulated data set
# q<-stats::quantile(test_stat_null,probs=1-signi)                               #Calculate 1-signi quantile of simulated distribution of the test statistic
# #decision<-as.numeric(teststat>q)                                       #Gives One if Hypothesis gets rejected, 0 otherwise
# f<-function(x,y){
#   r<-as.numeric(x>y)
#   return(r)
# }
# decision<-mapply(teststat,FUN=f,y=q)
# decision<-return(decision)
# }
# flares_arrivals <- flares %>% ctre() %>% thin(k = 700) %>% interarrival()
# p_value <- ML_LRT_boot(data=flares_arrivals,n=length(flares_arrivals))
```
